Further Modeling of $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak

### 2020-09-09 Papers/Paper2/Paper.tex
Paper2 ready for informal distruibution

\begin{abstract}
An improved resonant plasma response model that describes the interaction between a tokamak plasma and a resonant magnetic perturbation  (RMP) is developed. 
The model interpolates between the linear and the nonlinear response regimes, and takes into account the fact that the slip-frequency
is non-zero in the nonlinear regime. The improved model is incorporated into the {\tt EPEC} toroidal asymptotic matching code.
The modified {\tt EPEC} code is used to model RMP-induced edge-localized-modr mitigation/suppression in DIII-D H-mode discharge \#145380. 
Allowing for a finite slip-frequency (i.e., relaxing the so-called no-slip constraint) is found to slightly facilitate the locking of
driven magnetic island chains to the RMP. The previously obtained conclusion 
that the response of a typical H-mode tokamak plasma to an RMP cannot be accurately modeled by linear theory is confirmed. The previously obtained conclusion that the best agreement between theory and observations is achieved by assuming that the
natural frequencies of tearing modes, in the absence of the RMP, are determined by the local equilibrium ${\bf E}\times{\bf B}$ frequency is also confirmed.
\end{abstract}

@@

\section{Introduction}
In the context of tokamak plasmas, a resonant magnetic perturbation (RMP) refers to an externally
generated, (usually) static, helical magnetic perturbation that resonates at
one or more ``rational'' magnetic flux surfaces within the plasma. Resonant magnetic perturbations have been successfully
employed to compensate error fields (i.e., accidentally produced RMPs), and, thereby,  to prevent the associated formation of
locked modes, in a wide variety of different types of tokamak discharge. \cite{error1,error2,error3} RMPs have also been  successfully used to either  
mitigate or completely suppress  edge-localized-modes (ELMs) in H-mode tokamak discharges.\cite{elm1,elm2,elm3,elm4,elm5,elm6}

@@

The recently developed, toroidal, asymptotic matching code {\tt EPEC} has been used to simulate 
RMP-induced ELM mitigation/suppression in DIII-D H-mode plasmas.\cite{rftor,rftor1}
The original {\tt EPEC} model incorporates the no-slip constraint.
The first aim of this paper is to investigate 
what happens when the no-slip constraint is relaxed in the aforementioned {\tt EPEC} simulations.
 The second aim is to examine what happens when an improvement is made to the manner in which the resonant
plasma response component of the {\tt EPEC} model interpolates between the linear and the nonlinear regimes.

@@

\subsection{Plasma Equilibrium}
Consider a large aspect-ratio, low-$\beta$, tokamak plasma whose magnetic flux surfaces map out (almost)
concentric circles in the poloidal plane.
Such a plasma is well approximated as a periodic cylinder.  Suppose that the minor radius of the plasma
is $a$. Standard cylindrical coordinates ($r$, $\theta$, $z$) are adopted. The system is assumed to
be periodic in the $z$-direction, with periodicity length $2\pi\,R_0$, where $R_0\gg a$ is the simulated
plasma major radius. It is convenient to define the simulated toroidal angle $\phi=z/R_0$.

@@

\subsection{Plasma Response to RMP}
Consider the response of the plasma to an externally generated, static, helical, RMP. 
Suppose that the RMP has $m$ periods in the poloidal direction, and $n$ periods in the toroidal direction.
It is convenient to express the perturbed magnetic field and the perturbed plasma current density in
terms of a perturbed poloidal magnetic flux, $\psi(r,\theta,\phi,t)$. In fact, $\delta{\bf B} = \nabla\psi\times \nabla z$, 
and $\mu_0\,\delta {\bf j} = - \nabla^2\psi\,\nabla z$,
where $\psi(r,\theta,\phi,t) = \hat{\psi}(r,t)\,\exp[{\rm i}\,(m\,\theta-n\,\phi)]$, and $\hat{\psi}$ is real. This particular representation
is valid provided that $m/n\gg a/R_0$.\cite{rfa}

@@

Equation~(\ref{e5}) is usually coupled with the so-called ``no-slip constraint'':\,\cite{rfa}
\begin{equation}\label{e6}
\frac{d\varphi_s}{dt} = \omega_s.
\end{equation}
The reasoning behind the imposition of this constraint is that the plasma is trapped inside the
magnetic separatrix of the  island chain that forms in the vicinity of the rational surface,
which forces the chain to co-rotate with the local plasma flow. However, in reality, the plasma
is able to diffuse resistively across the separatrix to some extent. Hence, by analogy with Eqs.~(\ref{e3}), (\ref{e4}), and (\ref{e5}), and 
in accordance with Refs.~\onlinecite{slip1} and \onlinecite{slip2}, in this paper we shall modify Eq.~(\ref{e6}) such that
\begin{equation}\label{e7}
\frac{d\varphi_s}{dt} = \omega_s - \frac{E_{sv}}{\tau_R}\,\frac{r_s}{({\cal I}/2)\,W_s}\,\frac{|{\mit\Psi}_v|}{|{\mit\Psi}_s|}\,\sin(\varphi_s-\varphi_v).
\end{equation}
The final term on the right-hand side of the previous equation is the ``nonlinear slip-frequency''. In general,
we would expect this frequency to be relatively small (because it is generally the case that $|\omega_s|\,\tau_R\,(W_s/r_s)\gg 1$
for fully-developed magnetic islands in high-temperature tokamak plasmas\,\cite{rf1,slip2}). Nevertheless, the
nonlinear slip-frequency may be non-negligible for developing island chains, which is why we are including it in our analysis.
(Note that the nonlinear slip-frequency does not appear in rigorous island calculations, such as
that described in Ref.~\onlinecite{rfisland}, and references therein, because such calculations depend crucially on
complicated hierarchical ordering assumptions, according to which the nonlinear slip-frequency is negligible.)

@@

In the previous expressions, $R_0$ is the major radius of the magnetic axis, $r$ is a magnetic flux surface label with dimensions of length (which, roughly speaking, is the average minor radius of the
flux surface), $q(r)$ is the safety-factor profile, $s(r)=d\ln q/d\ln r$, 
$g(r)\,B_0$ is the toroidal magnetic field-strength at major radius $R_0$, $B_0$ is the vacuum
toroidal magnetic field-strength on the magnetic axis, $a$ is the value of $r$ on the last closed magnetic-flux surface, $r_k$ is the value of $r$ on the $k$th rational
surface (resonant with poloidal mode number $m_k$, and toroidal mode number $n$), $\hat{a}=a/R_0$, $\hat{r}=r/a$, and $\hat{r}_k=r_k/a$. 
Furthermore, $\tau_R(\hat{r})= \mu_0\,r^{\,2}\,\sigma_{ee}\,Q_{00}^{\,ee}$, where $\sigma_{ee}$ is the classical plasma parallel electrical conductivity, and
$Q_{00}^{\,ee}$ describes the corrections to this conductivity due to neoclassical effects and the presence of plasma impurities. Also, $\tau_A$ is a convenient
scale time, and $\hat{t}=t/\tau_A$. Finally, ${\mit\Psi}_k = B_0\,R_0\,\hat{\mit\Psi}_k\,{\rm e}^{-{\rm i}\,\varphi_k}$ is the reconnected magnetic flux
at the $k$th rational surface, ${\mit\chi}_k = B_0\,R_0\,\hat{\mit\chi}_k\,{\rm e}^{-{\rm i}\,\zeta_k}$ is a measure of the vacuum magnetic flux at the $k$th
rational surface (in fact, $\hat{E}_{kk}\,\chi_k$ is equivalent to $E_{sv}\,{\mit\Psi}_v$ in Sect.~\ref{s2}), and $E_{kk'} =\hat{E}_{kk'}\,{\rm e}^{-{\rm i}\,\xi_{kk'}}$ is the toroidal tearing stability matrix.\cite{am1} Here, $\hat{\mit\Psi}_k>0$, $\varphi_k$, $\hat{\chi}_k>0$,
$\zeta_k$, $\hat{E}_{kk'}>0$, and $\xi_{kk'}$ are all real quantities.

Equation~(\ref{e15}) represents a minor improvement to Eq.~(D1) of Ref.~\onlinecite{rftor} in which the interpolation between the
linear and nonlinear regimes is performed in a more systematic manner.

Equation~(D6) of Ref.~\onlinecite{rftor} is modified such that
\begin{align}\label{e19}
\left(\hat{W}_k +\hat{\delta}_k\right){\cal S}_k\left(\frac{d\varphi_k}{d\hat{t}}-\hat{\varpi}_k\right)&=
- \sum_{k'=1,K}\hat{E}_{kk'}\,\hat{\mit\Psi}_{k'}\,\sin(\varphi_k-\varphi_{k'}-\xi_{kk'})-\hat{E}_{kk}\,\hat{\chi}_k\,\sin(\varphi_k-\zeta_k),
\end{align}
where
\begin{equation}
\hat{\varpi}_k= \hat{\varpi}_{k\,0}+m_k\,{\mit\Delta}\hat{\mit\Omega}_\theta(\hat{r}_k,\hat{t})-n\,{\mit\Delta}\hat{\mit\Omega}_\phi(\hat{r}_k,\hat{t}),
\end{equation}
and $\hat{\varpi}_{k\,0}= \varpi_{k\,0}\,\tau_A$, $\hat{\mit\Omega}_\theta= {\mit\Delta}{\mit\Omega}_\theta\,\tau_A$, and $\hat{\mit\Omega}_\phi= {\mit\Delta}{\mit\Omega}_\phi\,\tau_A$. Here, $\varpi_{k\,0}$
is the so-called ``natural frequency'' at the $k$th rational surface; this quantity is defined as the helical phase velocity of a naturally unstable island chain, resonant at the surface, in the
absence of an RMP (or any other island chains). On the other hand,  ${\mit\Delta}{\mit\Omega}_\theta$ and ${\mit\Delta}{\mit\Omega}_\phi$ are the changes in the plasma
poloidal and toroidal velocity profiles, respectively, induced by the electromagnetic torques that develop in the vicinities of the various rational surfaces in the plasma.
Equation~(\ref{e19}) represents
a major improvement to Eq.~(D6) that takes into  account the finite slip-frequencies of the magnetic island chains induced at
the various rational surfaces in the plasma. In the absence of a finite slip-frequency, the right-hand side of Eq.~(\ref{e19}) would be zero.

@@

\section{EPEC Modeling of DIII-D Discharge \#145380}\label{s4}
\subsection{Introduction}
DIII-D discharge \#145380 is an ITER-Similar-Shape (ISS), ELMing, H-mode discharge, with a toroidal magnetic field $B_T=-1.9\,{\rm T}$, in which the
plasma current, $I_p$, is slowly ramped over a 2 second interval in order to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of this discharge. ${\mit\Psi}_N$ is the conventional normalized poloidal magnetic flux.
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression or mitigation are evident: the first (in which ELMs are mitigated, but not
entirely suppressed) extends from $t=2840$--$2980$ ms; the
second (in which ELM are entirely suppressed) extends from $t=3320$--$3560$ ms; and the third (in which ELMs are
entirely suppressed) extends from $t=3880$--$4200$ ms. The 
first, second, and third ELM-suppression/mitigation windows extend over the $\overline{q_{95}}$ ranges $3.88$--$3.94$, $3.62$--$3.73$,
and $3.33$--$3.48$, respectively. Here, $\overline{q_{95}}$ is least-squares linear fit to the variation of the safety-factor at the ${\mit\Psi}_N = 0.95$ flux surface with time.

@@

It can be seen, by a comparison of Figs.~\ref{fig7}--\ref{fig10}, that the linear calculation very significantly overestimates how easy it is for magnetic
island chains to lock to the RMP. The reason for this is that the linear slip-frequency is comparatively high, because the linear layer width is
comparatively narrow. The true slip-frequency becomes less than the linear slip-frequency as soon as the island width exceeds the
linear layer width. [See Eq.~(\ref{e10}).] The fact that that linear calculation is highly inaccurate confirms the conclusion reached in Ref.~\onlinecite{rf1}
that the response of a typical H-mode tokamak plasma to an RMP is not governed by linear theory, because the driven
island widths exceed the very narrow linear layer widths, even when driven reconnection is strongly suppressed by plasma flow.

Intriguingly, the composite linear/nonlinear model fairly accurately predicts the temporal extends of the three ELM mitigation/suppression windows (assuming that the
windows correspond to the formation of wide locked magnetic island chains at the top of the pedestal). 
However, the windows do not occur at the correct times. We conclude that Eq.~(\ref{eeb}) gives a more accurate prediction for the
natural frequencies, in the absence of the RMP, than Eq.~(\ref{ecomp}). However, the reason for this is, as yet, unclear.

\section{Discussion and Conclusions}
We have developed an improved resonant plasma response model that describes the interaction between a tokamak plasma and an RMP. 
The model interpolates between the linear and the nonlinear response regimes, and takes into account the fact that the slip-frequency
is non-zero in the nonlinear regime. Our improved model has been incorporated into the {\tt EPEC} toroidal asymptotic matching code.\cite{rftor1}
The moidifed {\tt EPEC} code is used to model RMP-induced ELM mitigation/suppression in DIII-D discharge \#145380.

@@

\bibitem{rftor1} R.~Fitzpatrick,  ``Modelling $q_{95}$ Windows for the Suppression of Edge Localized
Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak", submitted to Physics of Plasmas (2020).

@@

\begin{figure}
\includegraphics[height=6in]{fig2.pdf}
\caption{Top Panel: $n=3$ natural frequencies, in absence of RMP, as functions of the least-squares linear fit to $q_{95}$ versus time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by the ${\bf E}\times {\bf B}$
frequency.
Bottom Panel:  $n=3$ natural frequencies, in presence of RMP, as functions of time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by the ${\bf E}\times {\bf B}$
frequency. The red, green, blue, yellow, cyan, magenta, brown, pink,
purple, and orange  curves correspond to $m=5$, 6, 7, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression windows.} \label{fig2}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig3.pdf}
\caption{Top Panel: Full  $n=3$ vacuum island widths as functions of the least-squares linear fit to $q_{95}$ versus time 
in   DIII-D discharge \#145380.
Bottom Panel:  Full $n=3$ island widths as functions of time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by  the ${\bf E}\times {\bf B}$
frequency. The blue, yellow, cyan, magenta, brown, pink,
purple, and orange  areas correspond to $m=7$, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression/mitigation windows. 
The horizontal dotted lines indicate the top of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig3}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig4.pdf}
\caption{Top Panel: $n=3$ natural frequencies, in absence of RMP, as functions of the least-squares linear fit to $q_{95}$ versus time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by the ${\bf E}\times {\bf B}$
frequency.
Bottom Panel:  $n=3$ natural frequencies, in presence of RMP, as functions of time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by the ${\bf E}\times {\bf B}$
frequency. The red, green, blue, yellow, cyan, magenta, brown, pink,
purple, and orange  curves correspond to $m=5$, 6, 7, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression windows.} \label{fig4}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig5.pdf}
\caption{Top Panel: Full  $n=3$ vacuum island widths as functions of the least-squares linear fit to $q_{95}$ versus time 
in   DIII-D discharge \#145380.
Bottom Panel:  Full $n=3$ island widths as functions of time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by the ${\bf E}\times {\bf B}$
frequency. The blue, yellow, cyan, magenta, brown, pink,
purple, and orange  areas correspond to $m=7$, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression/mitigation windows. 
The horizontal dotted lines indicate the top of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig5}
\end{figure}

@@

\begin{figure}
\includegraphics[height=6in]{fig7.pdf}
\caption{Top Panel: $n=3$ natural frequencies, in absence of RMP, as functions of the least-squares linear fit to $q_{95}$ versus time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by linear theory.
Bottom Panel:  $n=3$ natural frequencies, in presence of RMP, as functions of time
in   DIII-D discharge \#145380, assuming that the natural frequency is determined by linear theory. The red, green, blue, yellow, cyan, magenta, brown, pink,
purple, and orange  curves correspond to $m=5$, 6, 7, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression windows.} \label{fig7}
\end{figure}

>>>>

\begin{abstract}
An improved resonant plasma response model that describes the interaction between a tokamak plasma and a resonant magnetic perturbation  (RMP) is developed. 
The model interpolates between the linear and the nonlinear response regimes, and takes into account the fact that the slip-frequency
is non-zero in the nonlinear regime. The improved model is incorporated into the {\tt EPEC} toroidal asymptotic matching code.
The modified {\tt EPEC} code is used to investigate RMP-induced edge-localized-mode (ELM) mitigation/suppression in DIII-D H-mode discharge \#145380. 
Allowing for a finite slip-frequency (i.e., relaxing the so-called no-slip constraint) is found to slightly facilitate the locking of
driven magnetic island chains to the RMP, and, hence, to slightly facilitate RMP-induced ELM mitigation/suppression.  The previously obtained conclusion 
that the response of a typical H-mode tokamak plasma to an RMP cannot be accurately modeled by linear theory is confirmed. The previously obtained conclusion that the best agreement between theory and observations is achieved by assuming that the
natural frequencies of tearing modes, in the absence of the RMP, are determined by the local equilibrium ${\bf E}\times{\bf B}$ frequency is also confirmed.
\end{abstract}

@@

\section{Introduction}
In the context of tokamak plasmas, a resonant magnetic perturbation (RMP) refers to an externally
generated, (usually) static, helical magnetic perturbation that resonates at
one or more ``rational'' magnetic flux-surfaces within the plasma. Resonant magnetic perturbations have been successfully
employed to compensate error fields (i.e., accidentally produced RMPs), and, thereby,  to prevent the associated formation of
locked modes, in a wide variety of different types of tokamak discharge. \cite{error1,error2,error3} RMPs have also been  successfully used to either  
mitigate or completely suppress  edge-localized-modes (ELMs) in H-mode tokamak discharges.\cite{elm1,elm2,elm3,elm4,elm5,elm6}

@@

The recently developed, toroidal, asymptotic matching code {\tt EPEC} (Extended Perturbed Equilibrium Code) has been used to simulate 
RMP-induced ELM mitigation/suppression in DIII-D H-mode plasmas.\cite{rftor,rftor1}
The original {\tt EPEC} model incorporates the no-slip constraint.
The first aim of this paper is to investigate 
what happens when the no-slip constraint is relaxed in the aforementioned {\tt EPEC} simulations.
 The second aim is to examine what happens when an improvement is made to the manner in which the resonant
plasma response component of the {\tt EPEC} model interpolates between the linear and the nonlinear regimes.

@@

\subsection{Plasma Equilibrium}
Consider a large aspect-ratio, low-$\beta$, tokamak plasma whose magnetic flux-surfaces map out (almost)
concentric circles in the poloidal plane.
Such a plasma is well approximated as a periodic cylinder.  Suppose that the minor radius of the plasma
is $a$. Standard cylindrical coordinates ($r$, $\theta$, $z$) are adopted. The system is assumed to
be periodic in the $z$-direction, with periodicity length $2\pi\,R_0$, where $R_0\gg a$ is the simulated
plasma major radius. It is convenient to define the simulated toroidal angle $\phi=z/R_0$.

@@

\subsection{Plasma Response to RMP}
Consider the response of the plasma to an externally generated, static, helical, RMP. 
Suppose that the RMP has $m$ periods in the poloidal direction, and $n$ periods in the toroidal direction.
It is convenient to express the perturbed magnetic field and the perturbed plasma current density in
terms of a perturbed poloidal magnetic flux (divided by $2\pi\,R_0$), $\psi(r,\theta,\phi,t)$. In fact, $\delta{\bf B} = \nabla\psi\times \nabla z$, 
and $\mu_0\,\delta {\bf j} = - \nabla^2\psi\,\nabla z$,
where $\psi(r,\theta,\phi,t) = \hat{\psi}(r,t)\,\exp[{\rm i}\,(m\,\theta-n\,\phi)]$, and $\hat{\psi}$ is real. This particular representation
is valid provided that $m/n\gg a/R_0$.\cite{rfa}

@@

Equation~(\ref{e5}) is usually coupled with the so-called ``no-slip constraint'':\,\cite{rfa}
\begin{equation}\label{e6}
\frac{d\varphi_s}{dt} = \omega_s.
\end{equation}
The reasoning behind the imposition of this constraint is that the plasma is trapped inside the
magnetic separatrix of the  island chain that forms in the vicinity of the rational surface,
which forces the chain to co-rotate with the local plasma flow. However, in reality, the plasma
is able to diffuse resistively across the separatrix to some extent. Hence, by analogy with Eqs.~(\ref{e3}), (\ref{e4}), and (\ref{e5}), and 
in accordance with Refs.~\onlinecite{slip1} and \onlinecite{slip2}, in this paper we shall modify Eq.~(\ref{e6}) such that
\begin{equation}\label{e7}
\frac{d\varphi_s}{dt} = \omega_s - \frac{E_{sv}}{\tau_R}\,\frac{r_s}{({\cal I}/2)\,W_s}\,\frac{|{\mit\Psi}_v|}{|{\mit\Psi}_s|}\,\sin(\varphi_s-\varphi_v).
\end{equation}
The final term on the right-hand side of the previous equation is the ``nonlinear slip-frequency''. In general,
we would expect this frequency to be relatively small (because it is usually the case that $|\omega_s|\,\tau_R\,(W_s/r_s)\gg 1$
for fully-developed magnetic islands in high-temperature tokamak plasmas\,\cite{rf1,slip2}). Nevertheless, the
nonlinear slip-frequency may be non-negligible for developing island chains, which is why we are including it in our analysis.
(Note that the nonlinear slip-frequency does not appear in rigorous island calculations, such as
that described in Ref.~\onlinecite{rfisland}, and references therein, because such calculations depend crucially on
complicated hierarchical ordering assumptions, according to which the nonlinear slip-frequency is negligible.)

@@

In the previous expressions, $R_0$ is the major radius of the magnetic axis, $r$ is a magnetic flux-surface label with dimensions of length (which, roughly speaking, is the average minor radius of the
flux-surface), $q(r)$ is the safety-factor profile, $s(r)=d\ln q/d\ln r$, 
$g(r)\,B_0$ is the toroidal magnetic field-strength at major radius $R_0$, $B_0$ is the vacuum
toroidal magnetic field-strength on the magnetic axis, $a$ is the value of $r$ on the last closed magnetic flux-surface, $r_k$ is the value of $r$ on the $k$th rational
surface (resonant with poloidal mode number $m_k$, and toroidal mode number $n$), $\hat{a}=a/R_0$, $\hat{r}=r/a$, and $\hat{r}_k=r_k/a$. 
Furthermore, $\tau_R(\hat{r})= \mu_0\,r^{\,2}\,\sigma_{ee}\,Q_{00}^{\,ee}$, where $\sigma_{ee}$ is the classical plasma parallel electrical conductivity, and
$Q_{00}^{\,ee}$ describes the corrections to this conductivity due to neoclassical effects and the presence of plasma impurities. Also, $\tau_A$ is a convenient
scale time, and $\hat{t}=t/\tau_A$. Finally, ${\mit\Psi}_k = B_0\,R_0\,\hat{\mit\Psi}_k\,{\rm e}^{-{\rm i}\,\varphi_k}$ is the reconnected magnetic flux
at the $k$th rational surface, ${\mit\chi}_k = B_0\,R_0\,\hat{\mit\chi}_k\,{\rm e}^{-{\rm i}\,\zeta_k}$ is a measure of the vacuum magnetic flux at the $k$th
rational surface (in fact, $\hat{E}_{kk}\,\chi_k$ is equivalent to $E_{sv}\,{\mit\Psi}_v$ in Sect.~\ref{s2}), and $E_{kk'} =\hat{E}_{kk'}\,{\rm e}^{-{\rm i}\,\xi_{kk'}}$ is the toroidal tearing stability matrix.\cite{am1} Here, $\hat{\mit\Psi}_k>0$, $\varphi_k$, $\hat{\chi}_k>0$,
$\zeta_k$, $\hat{E}_{kk'}>0$, and $\xi_{kk'}$ are all real quantities.

Equation~(\ref{e15}) represents a minor improvement to Eq.~(D1) of Ref.~\onlinecite{rftor} in which the interpolation between the
linear and nonlinear regimes is performed in a slightly more accurate manner.

Equation~(D6) of Ref.~\onlinecite{rftor} is modified such that
\begin{align}\label{e19}
\left(\hat{W}_k +\hat{\delta}_k\right){\cal S}_k\left(\frac{d\varphi_k}{d\hat{t}}-\hat{\varpi}_k\right)&=
- \sum_{k'=1,K}\hat{E}_{kk'}\,\hat{\mit\Psi}_{k'}\,\sin(\varphi_k-\varphi_{k'}-\xi_{kk'})-\hat{E}_{kk}\,\hat{\chi}_k\,\sin(\varphi_k-\zeta_k),
\end{align}
where
\begin{equation}
\hat{\varpi}_k= \hat{\varpi}_{k\,0}+m_k\,{\mit\Delta}\hat{\mit\Omega}_\theta(\hat{r}_k,\hat{t})-n\,{\mit\Delta}\hat{\mit\Omega}_\phi(\hat{r}_k,\hat{t}),
\end{equation}
and $\hat{\varpi}_{k\,0}= \varpi_{k\,0}\,\tau_A$, $\hat{\mit\Omega}_\theta= {\mit\Delta}{\mit\Omega}_\theta\,\tau_A$, and $\hat{\mit\Omega}_\phi= {\mit\Delta}{\mit\Omega}_\phi\,\tau_A$. Here, $\varpi_{k\,0}$
is the so-called ``natural frequency'' at the $k$th rational surface; this quantity is defined as the helical phase velocity of a naturally unstable island chain, resonant at the surface, in the
absence of an RMP (or any other island chains). On the other hand,  ${\mit\Delta}{\mit\Omega}_\theta$ and ${\mit\Delta}{\mit\Omega}_\phi$ are the changes in the plasma
poloidal and toroidal velocity profiles, respectively, induced by the electromagnetic torques that develop in the vicinities of the various rational surfaces in the plasma in response to the RMP.
Equation~(\ref{e19}) represents
a major improvement to Eq.~(D6) of Ref.~\onlinecite{rftor} that takes into  account the finite slip-frequencies of the magnetic island chains induced at
the various rational surfaces in the plasma. In the absence of a finite slip-frequency, the right-hand side of Eq.~(\ref{e19}) would be zero.

@@

\section{{\tt EPEC} Modeling of DIII-D Discharge \#145380}\label{s4}
\subsection{Introduction}
DIII-D discharge \#145380 is an ITER-Similar-Shape (ISS), ELMing, H-mode discharge, with a toroidal magnetic field $B_T=-1.9\,{\rm T}$, in which the
plasma current, $I_p$, is slowly ramped over a 2 second interval in order to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of this discharge. ${\mit\Psi}_N$ is the conventional normalized poloidal magnetic flux.
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression or mitigation are evident: the first (in which ELMs are mitigated, but not
entirely suppressed) extends from $t=2840$--$2980$ ms; the
second (in which ELM are entirely suppressed) extends from $t=3320$--$3560$ ms; and the third (in which ELMs are
entirely suppressed) extends from $t=3880$--$4200$ ms. The 
first, second, and third ELM-suppression/mitigation windows extend over the $\overline{q_{95}}$ ranges $3.88$--$3.94$, $3.62$--$3.73$,
and $3.33$--$3.48$, respectively. Here, $\overline{q_{95}}$ is least-squares linear fit to the variation of the safety-factor at the ${\mit\Psi}_N = 0.95$ flux-surface with time.

@@

It can be seen, by a comparison of Figs.~\ref{fig7}--\ref{fig10}, that the linear calculation very significantly overestimates how easy it is for magnetic
island chains to lock to the RMP. The reason for this is that the linear slip-frequency is comparatively high, because the linear layer width is
comparatively narrow. The true slip-frequency becomes less than the linear slip-frequency as soon as the island width exceeds the
linear layer width. [See Eq.~(\ref{e10}).] The fact that the linear calculation is highly inaccurate confirms the conclusion reached in Ref.~\onlinecite{rf1}
that the response of a typical H-mode tokamak plasma to an RMP is not governed by linear theory, because the driven
island widths exceed the very narrow linear layer widths, even when driven reconnection is strongly suppressed by plasma flow.

Intriguingly, the composite linear/nonlinear model fairly accurately predicts the temporal extends of the three ELM mitigation/suppression windows (assuming that the
windows correspond to the formation of wide locked magnetic island chains at the top of the pedestal). 
However, the windows do not occur in the correct time intervals. We conclude that Eq.~(\ref{eeb}) gives a more accurate prediction for the
natural frequencies, in the absence of the RMP, than Eq.~(\ref{ecomp}). However, the reason for this is, as yet, unclear.

\section{Summary}
We have developed an improved resonant plasma response model that describes the interaction between a tokamak plasma and an RMP. 
The model interpolates between the linear and the nonlinear response regimes, and takes into account the fact that the slip-frequency
is non-zero in the nonlinear regime. Our improved model has been incorporated into the {\tt EPEC} toroidal asymptotic matching code.\cite{rftor1}
The modified {\tt EPEC} code is used to model RMP-induced ELM mitigation/suppression in DIII-D H-mode discharge \#145380.

@@

\bibitem{rftor1} R.~Fitzpatrick,  ``Modeling $q_{95}$ Windows for the Suppression of Edge Localized
Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak", submitted to Physics of Plasmas (2020).

@@

\begin{figure}
\includegraphics[height=6in]{fig2.pdf}
\caption{Composite linear/nonlinear calculation with the no-slip constraint imposed at all rational surfaces, and the natural frequencies
in the absence of the RMP determined by the local  ${\bf E}\times {\bf B}$
frequency.
Top Panel: $n=3$ natural frequencies, in absence of RMP, as functions of the least-squares linear fit to $q_{95}$ versus time
in   DIII-D discharge \#145380. Bottom Panel:  $n=3$ natural frequencies, in presence of RMP, as functions of time
in   DIII-D discharge \#145380. The red, green, blue, yellow, cyan, magenta, brown, pink,
purple, and orange  curves correspond to $m=5$, 6, 7, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression windows.} \label{fig2}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig3.pdf}
\caption{Composite linear/nonlinear calculation with the no-slip constraint imposed at all rational surfaces, and the natural frequencies
in the absence of the RMP determined by the local  ${\bf E}\times {\bf B}$
frequency. Top Panel: Full  $n=3$ vacuum island widths as functions of the least-squares linear fit to $q_{95}$ versus time 
in   DIII-D discharge \#145380.
Bottom Panel:  Full $n=3$ island widths as functions of time
in   DIII-D discharge \#145380. The blue, yellow, cyan, magenta, brown, pink,
purple, and orange  areas correspond to $m=7$, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression/mitigation windows. 
The horizontal dotted lines indicate the top of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig3}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig4.pdf}
\caption{Composite linear/nonlinear calculation with the no-slip constraint relaxed at all rational surfaces, and the natural frequencies
in the absence of the RMP determined by the local  ${\bf E}\times {\bf B}$
frequency.
Top Panel: $n=3$ natural frequencies, in absence of RMP, as functions of the least-squares linear fit to $q_{95}$ versus time
in   DIII-D discharge \#145380.
Bottom Panel:  $n=3$ natural frequencies, in presence of RMP, as functions of time
in   DIII-D discharge \#145380. The red, green, blue, yellow, cyan, magenta, brown, pink,
purple, and orange  curves correspond to $m=5$, 6, 7, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression windows.} \label{fig4}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig5.pdf}
\caption{Composite linear/nonlinear calculation with the no-slip constraint relaxed at all rational surfaces, and the natural frequencies
in the absence of the RMP determined by the local  ${\bf E}\times {\bf B}$
frequency. Top Panel: Full  $n=3$ vacuum island widths as functions of the least-squares linear fit to $q_{95}$ versus time 
in   DIII-D discharge \#145380.
Bottom Panel:  Full $n=3$ island widths as functions of time
in   DIII-D discharge \#145380. The blue, yellow, cyan, magenta, brown, pink,
purple, and orange  areas correspond to $m=7$, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression/mitigation windows. 
The horizontal dotted lines indicate the top of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig5}
\end{figure}

@@

\begin{figure}
\includegraphics[height=6in]{fig7.pdf}
\caption{Linear calculation with the no-slip constraint relaxed at all rational surfaces, and  the natural frequencies
in the absence of the RMP determined by linear theory. 
 Top Panel: $n=3$ natural frequencies, in absence of RMP, as functions of the least-squares linear fit to $q_{95}$ versus time
in   DIII-D discharge \#145380.
Bottom Panel:  $n=3$ natural frequencies, in presence of RMP, as functions of time
in   DIII-D discharge \#145380. The red, green, blue, yellow, cyan, magenta, brown, pink,
purple, and orange  curves correspond to $m=5$, 6, 7, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression windows.} \label{fig7}
\end{figure}

\begin{figure}
\includegraphics[height=6in]{fig8.pdf}
\caption{Linear calculation with the no-slip constraint relaxed at all rational surfaces, and  the natural frequencies
in the absence of the RMP determined by linear theory. Top Panel: Full  $n=3$ vacuum island widths as functions of the least-squares linear fit to $q_{95}$ versus time 
in   DIII-D discharge \#145380.
Bottom Panel:  Full $n=3$ island widths as functions of time
in   DIII-D discharge \#145380. The blue, yellow, cyan, magenta, brown, pink,
purple, and orange  areas correspond to $m=7$, 8, 9, 10, 11, 12, 13, and 14, respectively. The yellow vertical bands indicate the ELM-suppression/mitigation windows. 
The horizontal dotted lines indicate the top of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig8}
\end{figure}
